---
title: "Marijuana and Alcohol"
author: "Jiaye Wang (UID 906147831)"
date: "2026-02-27"
output: pdf_document
---

# Setup

```{r}
# Packages not included with a standard R install (uncomment to install once)
#install.packages("fixest")
#install.packages("gt")
#install.packages("broom")
#install.packages("purrr")
#install.packages("modelsummary")

library(readxl)
library(ggplot2)
library(tidyr)
library(dplyr)
library(fixest)
library(modelsummary)
library(broom)
library(purrr)

# Sheet "Graph 1": Year, CA, ControlledStateAverage
# Sheet "Graph 2": Year, CA, one column per control state (18), Average
data_file <- "Alcohol Consumption Data.xlsx"
```

# 1. Data Preparation

Two versions of the data are used:

* `df_trend` (wide, one row per year): California vs. the control-state average.
  Used for the trend and counterfactual plots.
* `df_panel` (long, one row per state-year): California plus the 18 control states.
  Used for the summary statistics, event study, and DiD regression.

```{r}
# Wide trend data: CA vs. control-state average (sheet "Graph 1")
df_trend <- read_excel(data_file, sheet = "Graph 1")

# Long state-year panel (CA from "Graph 1", controls from "Graph 2")
# CA in panel format
ca <- df_trend |>
  select(Year, CA) |>
  rename(year = Year, y = CA) |>
  mutate(state = "CA")

# Control states in panel format
control <- read_excel(data_file, sheet = "Graph 2") |>
  select(-Average) |> # drop the "Average" column
  rename(year = Year) |>
  pivot_longer(cols = -year, names_to = "state", values_to = "y")

# Combine CA with controls and build treatment indicators
df_panel <- bind_rows(ca, control) |>
  mutate(treated = (state == "CA"),                        # treatment group
         post = (year >= 2018),                            # post-legalization (retail sales began 1/1/2018)
         did = treated * post,                             # CA x Post interaction
         treat_year = ifelse(state == "CA", 2018, 0)       # 0 = never-treated controls
  )
```

# 2. Summary Statistics (Section 3.2, Table 1)

```{r}
df_sum <- df_panel |>
  mutate(
    # grouping variables
    group = ifelse(state == "CA", "California", "Control States"),
    period = ifelse(year < 2018, "Pre-2018", "Post-2018")
  )

# Table 1: Summary Statistics: Alcohol Consumption by Group
datasummary(
  y ~ group * (Mean + SD + N),
  data = df_sum,
  title = "Summary Statistics: Alcohol Consumption by Group",
  output = "gt"
)
```

# 3. No Anticipation Assumption (Paper Section 4.1)

Event study using the Sun & Abraham (`sunab`) estimator with state and year fixed effects. 2017 (t = -1) is the reference year; pre-period coefficients near zero support no anticipation.

```{r}
es <- feols(y ~ sunab(treat_year, year) | state + year, data = df_panel, cluster = "state")
iplot(es,
      main = "Dynamic Impact of CA Recreational Marijuana Legalization",
      xlab = "Years Relative to Legalization",
      ylab = "Estimated Effect on Alcohol Consumption")
```

# 4. Parallel Trend Assumption (Section 4.2)

### 4a. California vs. control-state average

```{r}
ggplot(df_trend, aes(x = Year)) +
  # California trend line with dots
  geom_line(aes(y = CA, color = "CA"), linewidth = 1) +
  geom_point(aes(y = CA, color = "CA"), size = 2) +
  # Control trend line with dots
  geom_line(aes(y = ControlledStateAverage, color = "Controlled State Average"), linewidth = 1) +
  geom_point(aes(y = ControlledStateAverage, color = "Controlled State Average"), size = 2) +
  # Treatment start
  geom_vline(xintercept = 2018, linetype = "dashed") +
  # Labels
  labs(
    title = "Legal Recreational Marijuana's Effects on Alcohol Consumption",
    subtitle = "Marijuana Introduction in CA = 2018",
    x = "Year",
    y = "Alcohol Consumption (Gallons of Ethanol Per Capita)",
    color = "State"
  ) +
  theme_minimal()
```

### 4b. Counterfactual trend for California

The counterfactual is the control-state average shifted up by the mean pre-2018 gap between California and the controls, i.e. what California would look like without legalization if parallel trends hold.

```{r}
# Mean pre-treatment gap (CA - control)
pre_gap <- df_trend |>
  filter(Year < 2018) |>
  summarise(gap = mean(CA - ControlledStateAverage)) |>
  pull(gap)

# Counterfactual: CA without legalization of recreational marijuana
df_counterfactual <- df_trend |>
  mutate(CA_counterfactual = ControlledStateAverage + pre_gap)

ggplot(df_counterfactual, aes(x = Year)) +
  # Three lines: actual treated, counterfactual, and control
  geom_line(aes(y = CA, color = "CA (Actual)"), linewidth = 1) +
  geom_line(aes(y = ControlledStateAverage, color = "Control Avg"), linewidth = 1) +
  geom_line(aes(y = CA_counterfactual, color = "CA Counterfactual"),
            linetype = "dashed", linewidth = 1) +
  # Treatment start
  geom_vline(xintercept = 2018, linetype = "dashed") +
  # Labels
  labs(
    title = "Counterfactual Parallel Trend for CA",
    subtitle = "Dashed line shows expected CA alcohol consumption without legalization",
    y = "Alcohol Consumption (Gallons of Ethanol Per Capita)",
    x = "Year",
  ) +
  theme_minimal()
```

# 5. Difference-in-Differences Estimate (Paper Section 4.3, Table 2)

Model: $Y_{st} = \alpha + \beta (CA_s \times Post_t) + \gamma_s + \delta_t + \epsilon_{st}$, with state and year fixed effects and standard errors clustered by state.

```{r}
m_did <- feols(y ~ did | state + year, data = df_panel, cluster = "state")
modelsummary(m_did,
             coef_map = c("did" = "California × Post (DiD Effect)"),
             statistic = "({std.error})",
             gof_map = c("nobs", "r.squared"), # omit AIC, BIC, etc.
             title = "Difference-in-Differences Estimates of the Effect of Recreational Marijuana Legalization on Alcohol Consumption in CA"
)
```

# 6. Robustness Check (Paper Section 4.3)

Re-estimates the DiD coefficient (a) on the full sample (baseline), (b) excluding the COVID years 2020–2022, and (c) dropping each control state one at a time, then plots each estimate with its 95% confidence interval.

```{r}
# Control states and full state list
control_states <- c("AL","AR","GA","FL","ID","IN","IA","KS","KY","LA",
                    "MS","NE","NC","SC","TN","TX","WI","WY")
all_states <- c("CA", control_states)

# Long panel built directly from sheet "Graph 2"
df_robust <- read_excel(data_file, sheet = "Graph 2") |>
  select(Year, all_of(all_states)) |>
  pivot_longer(cols = -Year, names_to = "state", values_to = "alcohol") |>
  mutate(
    Post = if_else(Year >= 2018, 1, 0),
    CA = if_else(state == "CA", 1, 0)
  )

# Run the DiD on a given sample and return the CA:Post coefficient and 95% CI
run_did <- function(data, label) {
  m <- lm(alcohol ~ CA*Post + factor(state) + factor(Year), data = data)
  out <- tidy(m, conf.int = TRUE) |>
    filter(term == "CA:Post") |>
    transmute(
      spec = label,
      beta = estimate,
      se = std.error,
      p_value = p.value,
      ci_low = conf.low,
      ci_high = conf.high
    )
  out
}

#(a)Baseline: all states, all years
baseline_results <- run_did(
  df_robust |>
    filter(state %in% c("CA", control_states)),
  "Baseline"
)

#(b) Exclude COVID years
covid_results <- run_did(
  df_robust |>
    filter(state %in% c("CA", control_states), Year <= 2019),
  "Exclude 2020–2022"
)

#(c) Drop one control state at a time
control_results <- map_dfr(control_states, function(st) {
  run_did(
    df_robust |>
      filter(state %in% c("CA", setdiff(control_states, st))),
    paste("Drop", st)
  )
})

results_df <- bind_rows(baseline_results, covid_results, control_results)

# Keep specs in the order above (top to bottom on the plot)
results_df <- results_df |>
  mutate(spec = factor(spec, levels = rev(spec)))

# Robustness plot
ggplot(results_df, aes(x = beta, y = spec)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.2) +
  geom_point(size = 2.5) +
  labs(
    title = "Robustness of DiD Estimate",
    subtitle = "Marijuana legalization and alcohol consumption in CA",
    x = "Estimated CA × Post Effect",
    y = ""
  ) +
  theme_minimal()
```
